Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-06-20 CGP Journal/main.tex
Added conclusion

\subsection{Source of High Variance}
TODO Look for general indicators of high evaluation runs.

TODO Look for footprint moving.

TODO Look for active node clustering.

TODO Look into how often children replace parents.

@@

\section{Conclusion}
\label{sec:conclusion}
TODO The conclusion goes here.

>>>>

@@

\section{Conclusion}
\label{sec:conclusion}
Even with CGP's relatively simple evolutionary mechanisms, fully understanding
the effects those mechanisms have on search takes concerted effort.  By examining
the effectiveness as well as the actual behavior of traditional CGP and
a handful of variants we have helped advance this understanding.

Generally, CGP appears to require the least number of evaluations when the genome
size far exceeds what is necessary for the problem at hand.  When properly tuned,
almost all variants had only a tiny fraction of the genome active in their final
solution.  Furthermore, a significant portion of these active nodes were found
to be redundant, removable using a very simple reduction algorithm.  The only
variant that preferred more reasonable genome sizes were those using \emph{Single},
especially when paired with \emph{Reorder} or \emph{DAG}.  Yet we theorize this may
have more to do with \emph{Single} scaling poorly with the number of active nodes
than with it overcoming whatever causes traditional CGP to prefer enormous genomes.
While \emph{Reorder} and \emph{DAG} were able to use smaller genomes than \emph{Normal}
when paired with \emph{Single}, this decrease in size was not observed in any other
pairings.  As such, the suggestion in~\cite{goldman:2013:ordering} that these
techniques generally reduce genome size should be discarded.  Instead it is likely
more correct to say that \emph{Reorder} and \emph{DAG} are less detrimentally effected
by reducing the genome size.

Compared on median evaluations to success, \emph{Normal} and \emph{Reorder} are
likely to be equally good for simple problems, with some suggestion that \emph{Reorder}
performs better as problems become more difficult.  In general \emph{DAG} was found
to be unnecessarily time consuming, as it required higher algorithmic complexity
without any performance gain.  As such we would suggest \emph{Reorder} for general
future CGP use.

In line with the results from~\cite{goldman:2013:cgpwaste}, \emph{Skip} and
\emph{Accumulate} continued to show no significant difference in effectiveness
or behavior.  As \emph{Skip} is simpler to understand and implement, we suggest
\emph{Accumulate} be ignored in all future CGP applications.  \emph{Single}'s
behavior continues to be enough different from \emph{Skip} to warrant further
study, especially with problems requiring higher numbers of active genes.  In
general though, if evaluation time is significantly impacted by the number of
active nodes or the total size of the genome, we suggest \emph{Single}.  While
mutation rate appears relatively variant and problem independent (all of our tuning
resulted in 0.002 being optimal or close to it), \emph{Single} still has the
advantage of not requiring this parameter at all.

As to actual behavior, CGP genomes include a surprising amount of redundant and
unused nodes.  As theoretically predicted in~\cite{goldman:2013:ordering},
\emph{Normal} ordering leaves large sections of the genome completely unused
while \emph{Reorder} and \emph{DAG} are more likely to utilize the whole genome.
Further analysis showed that CGP, at least with our function set, produced a large
number of constant nodes, such that their output is independent of problem
inputs.  As the binary domain only has two constants (always true and always false),
each genome at most should require two constant nodes.  Even so these behaviors
were found duplicated in both the active and inactive portions of final solutions
for all algorithm variants.  Also common was duplication of node behavior, even
though CGP's encoding does not require any duplication to produce any value.  Yet
attempting to prevent or remove this behavioral duplication may hinder CGP's evolutionary
success, as it will limit neutral drift.

CGP's use of neutrality has become one of its cornerstone arguments for its success.
Yet the reason for neutrality's importance appears to be different from previous
arguments.  In our testing of traditional CGP we found large sections of the
genome were never used by any ancestor of the final solution.  Furthermore,
offspring almost never include active nodes that were inactive in their direct
parent, but active in a previous ancestor.  Even when reactivation does happen,
the node has almost always changed behavior significantly from when it was active.
As such we suggest the advantages of neutrality in regards to inactive nodes is
not that it allows the storage of previously useful structures, but that it allows
inactive nodes to be constantly changed facilitating mutational inclusion of what
are effectively random nodes.  Under this theory, CGP would likely benefit from randomly resetting
all inactive nodes just before they become active.  Note that with \emph{Reorder}
and \emph{DAG} these conclusions do not hold as a reasonable amount of node
reactivation occurs.

While this analysis was limited to boolean problems by necessity of some of
the analysis methods, most of the conclusions have no obvious reason to be
limited to the boolean domain.  Still, applying \emph{footprint} analysis
to detect changes in behavior and behavior duplication, along with looking at
never active nodes and tuned parameter settings, on other domains will help
solidify our conclusions.
